---
title: "Claims Frequency GLM Models Exploration"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
  pdf_document:
    latex_engine: xelatex
  word_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries Load

This section loads all packages required for the claims frequency analysis. Core modelling dependencies include *MASS* for the Negative Binomial GLM, *recipes* and *rsample* for reproducible preprocessing and cross-validation, and *survival* and *pROC* for shared evaluation infrastructure used across the broader feature selection exercise. All diagnostic visualisations are produced using *ggplot2*, with *tidyr* supporting the decile-level reshaping required by the lift charts.

```{r, include=FALSE}
library(here)
library(dplyr)
library(rsample)
library(recipes)
library(MASS)
library(pROC)
library(survival)
library(knitr)
library(ggplot2)
library(gridExtra)
library(nortest)
library(goftest)
library(corrplot)
```

# Data Exploration

First and foremost, we check the distribution of the claims count variable and the offset variable, `exposure_years`. A cap of 1 is imposed on the latter given the negligible number of policies exceeding a full-year duration and the practical convention that motor insurance policies are written annually.

```{r, echo=FALSE}
basic_glm_data <- read.csv(
  here("Feature_Selection_Four_Wheels_GLM_R.csv")
)

hist(
  basic_glm_data$n_claims_year,
  main   = "Claims Count Distribution",
  xlab   = "n_claims_year",
  col    = "blue",
  border = "white"
)

par(mfrow = c(1, 2))

hist(
  basic_glm_data$exposure_years,
  main   = "Original Exposure Years",
  xlab   = "exposure_years",
  col    = "skyblue",
  border = "white"
)

hist(
  pmin(basic_glm_data$exposure_years, 1),
  main   = "Exposure Years (Capped at 1)",
  xlab   = "exposure_years",
  col    = "tomato",
  border = "white"
)

basic_glm_data$exposure_years <- pmin(basic_glm_data$exposure_years, 1)
```


# Data Pre-processing

A client-stratified 80:20 train-test split is applied, where stratification is defined jointly by whether the client has ever lapsed, whether they have ever claimed, and the number of policies held. This ensures both partitions are representative of the full client population and that no client's policies appear in both sets. A grouped 5-fold cross-validation structure is subsequently constructed over the training set using client identifiers as the grouping variable. All categorical predictors are encoded using treatment-coded dummy variables derived from a fixed schema, with the encoding recipe re-learned on each fold's analysis partition independently to prevent leakage.

```{r, include=FALSE}

# --- Variable taxonomy -------------------------------------------------------

one_hot_vars <- c(
  "area_label", "channel_label", "payment_label", "second_driver_label",
  "type_fuel", "seniority_category", "start_season", "start_day_of_week"
)

ord_vars <- c(
  "age_category", "age_when_licensed_band", "driving_exp_category",
  "vehicle_age_category", "power_band", "cylinder_band", "value_band",
  "weight_band", "length_imputed_band", "n_doors_band",
  # "claims_history_band"       # dropped: near-perfect collinearity
  # "r_claims_history_band"     # dropped: near-perfect collinearity
  "avg_claims_per_year_history_band", "cross_sell_ratio_band",
  "policy_retention_rate_band", "risk_score_band"
)

binary_vars <- c(
  "growing_portfolio", "had_previous_lapse", "multi_policy_holder",
  "multi_product_holder", "start_weekend"
)

cat(sprintf(
  "Predictor features: %d total  (%d nominal OHE | %d ordinal | %d binary)\n",
  length(one_hot_vars) + length(ord_vars) + length(binary_vars),
  length(one_hot_vars), length(ord_vars), length(binary_vars)
))


# --- Client-stratified 80:20 split -------------------------------------------

set.seed(42)

client_strat <- basic_glm_data %>%
  group_by(id) %>%
  summarise(
    ever_lapsed  = as.integer(any(lapse_flag == 1,   na.rm = TRUE)),
    ever_claimed = as.integer(any(n_claims_year > 0, na.rm = TRUE)),
    n_policies   = n(),
    .groups = "drop"
  ) %>%
  mutate(
    policy_tier = case_when(
      n_policies == 1 ~ "1",
      n_policies == 2 ~ "2",
      TRUE            ~ "3+"
    ),
    strata = paste(ever_lapsed, ever_claimed, policy_tier, sep = "_")
  )

rare_strata  <- names(table(client_strat$strata)[table(client_strat$strata) == 1])
client_strat <- client_strat %>%
  mutate(strata = if_else(strata %in% rare_strata, "other", strata))

split_obj <- initial_split(client_strat, prop = 0.8, strata = strata)
train_ids  <- training(split_obj)$id
test_ids   <- testing(split_obj)$id

train_data <- basic_glm_data %>% filter(id %in% train_ids)
test_data  <- basic_glm_data %>% filter(id %in% test_ids)

cat(sprintf(
  "Train: %d policies | %d clients\nTest : %d policies | %d clients\n",
  nrow(train_data), n_distinct(train_data$id),
  nrow(test_data),  n_distinct(test_data$id)
))

stopifnot(length(intersect(unique(train_data$id), unique(test_data$id))) == 0)
cat("Client leakage check: PASSED\n")


# --- 5-fold grouped cross-validation ----------------------------------------

set.seed(42)
cv_folds <- group_vfold_cv(train_data, group = id, v = 5)

leakage_free <- vapply(cv_folds$splits, function(sp) {
  length(intersect(unique(analysis(sp)$id), unique(assessment(sp)$id))) == 0L
}, logical(1))

cat("Number of CV folds:", nrow(cv_folds), "\n")
cat("Client-leakage-free across all folds:", all(leakage_free), "\n")


# --- Encoding helpers --------------------------------------------------------

ordinal_levels <- list(
  age_category                     = c("18-25","26-35","36-45","46-55","56-65","65+"),
  age_when_licensed_band           = c("<=25",">25"),
  driving_exp_category             = c("<=5yrs","6-10yrs","11-20yrs","20+yrs"),
  vehicle_age_category             = c("0-3yrs","4-7yrs","8-12yrs","12+yrs"),
  power_band                       = c("<100",">100"),
  cylinder_band                    = c("0-1400","1400-1600","1600-2000","2000+"),
  value_band                       = c("0-17500","17500-25000","25000-32500","32500+"),
  weight_band                      = c("0-1100","1100-1500","1500+"),
  length_imputed_band              = c("<4","4-4.5",">4.5"),
  n_doors_band                     = c("<=4",">4"),
  claims_history_band              = c("zero","one","two_to_three","four_to_five","six_plus"),
  r_claims_history_band            = c("0","0-0.35","0.35-0.7","0.7-1.5","1.5+"),
  avg_claims_per_year_history_band = c("0","0-1",">1"),
  cross_sell_ratio_band            = c("<0.75",">=0.75"),
  policy_retention_rate_band       = c("<1","1"),
  risk_score_band                  = c("0","0-0.01","0.01-0.15","0.15-0.2","0.2+"),
  premium_band                     = c("0-250","250-300","300-350","350-400","400+"),
  premium_vs_median_band           = c("-15%","-15% to 0%","0% to 15%","15% to 30%","30%+")
)

recode_ordinals <- function(df) {
  for (v in intersect(names(ordinal_levels), names(df)))
    df[[v]] <- factor(df[[v]], levels = ordinal_levels[[v]], ordered = FALSE)
  df
}

always_exclude  <- c(
  "id", "id2", "n_claims_year", "avg_claim_cost",
  "lapse_flag", "days_to_event", "exposure_years",
  "r_claims_history_band", "claims_history_band"
)
premium_exclude <- c("premium_band", "premium_vs_median_band")

get_predictor_cols <- function(data, target) {
  extra_drop <- if (target %in% c("n_claims_year", "avg_claim_cost"))
    premium_exclude else character(0)
  setdiff(names(data), unique(c(always_exclude, extra_drop, target)))
}

encode_predictors <- function(train_df, test_df, target) {
  pred_cols <- get_predictor_cols(train_df, target)
  x_train   <- recode_ordinals(train_df[, pred_cols, drop = FALSE])
  x_test    <- recode_ordinals(test_df[, pred_cols, drop = FALSE])

  rec <- recipe(~ ., data = x_train) %>%
    step_novel(all_nominal_predictors()) %>%
    step_unknown(all_nominal_predictors()) %>%
    step_dummy(all_nominal_predictors(), one_hot = FALSE) %>%
    step_zv(all_predictors())

  prep_rec <- prep(rec, training = x_train, retain = TRUE)
  list(train  = bake(prep_rec, new_data = NULL),
       test   = bake(prep_rec, new_data = x_test),
       recipe = prep_rec)
}

strip_suffix <- function(dummy_names, known_vars = NULL) {
  if (is.null(known_vars)) return(sub("_[^_]+$", "", dummy_names))
  sapply(dummy_names, function(nm) {
    candidates <- known_vars[startsWith(nm, paste0(known_vars, "_"))]
    if (length(candidates) > 0) candidates[which.max(nchar(candidates))] else nm
  }, USE.NAMES = FALSE)
}


# --- Evaluation helpers ------------------------------------------------------

rmse <- function(obs, pred)  sqrt(mean((obs - pred)^2,  na.rm = TRUE))
mae  <- function(obs, pred)  mean(abs(obs  - pred),     na.rm = TRUE)

pseudo_r2_glm <- function(fit) 1 - fit$deviance / fit$null.deviance

cv_se <- function(x) sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))

pearson_gof <- function(fit) {
  X2 <- sum(residuals(fit, type = "pearson")^2)
  df <- df.residual(fit)
  list(
    statistic  = X2,
    df         = df,
    dispersion = X2 / df,
    p_value    = pchisq(X2, df = df, lower.tail = FALSE)
  )
}

ks_gof <- function(fit) {
  dr <- residuals(fit, type = "deviance")
  ks.test(dr, "pnorm", mean(dr), sd(dr))
}

ad_gof <- function(fit) {
  dr <- residuals(fit, type = "deviance")
  nortest::ad.test(dr)
}

cvm_gof <- function(fit) {
  dr <- residuals(fit, type = "deviance")
  goftest::cvm.test(dr, "pnorm", mean(dr), sd(dr))
}

run_wald_test <- function(fit, label) {
  b   <- coef(fit)
  V   <- vcov(fit)
  idx <- which(!grepl("^\\(Intercept\\)$", names(b)))
  k   <- length(idx)
  W   <- as.numeric(t(b[idx]) %*% solve(V[idx, idx]) %*% b[idx])
  w_ratio <- W / k
  p   <- pchisq(W, df = k, lower.tail = FALSE)
  sig <- if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""

  cat(sprintf(
    "\n[%s]\n  Wald chi2(%d) = %.4f   Ratio = %.2f   p-value = %.6f  %s\n",
    label, k, W, w_ratio, p, sig
  ))

  invisible(data.frame(
    Component  = label,
    Wald_Chi2  = round(W, 4),
    df         = k,
    Wald_Ratio = round(w_ratio, 2),
    p_value    = round(p, 6),
    stringsAsFactors = FALSE
  ))
}

export_glm_summary <- function(fit, prefix) {
  coef_df <- as.data.frame(summary(fit)$coefficients)
  coef_df <- data.frame(
    term      = rownames(coef_df),
    estimate  = coef_df[, 1], std_error = coef_df[, 2],
    statistic = coef_df[, 3], p_value   = coef_df[, 4],
    stringsAsFactors = FALSE
  )
  write.csv(coef_df, here(paste0(prefix, "_coefficients.csv")), row.names = FALSE)
  writeLines(capture.output(print(summary(fit))),
             here(paste0(prefix, "_summary.txt")))
  cat(sprintf("Exported: %s_coefficients.csv and %s_summary.txt\n", prefix, prefix))
}


# --- Shared encoding matrices ------------------------------------------------

enc_all <- encode_predictors(train_data, test_data, "n_claims_year")
X_train <- enc_all$train
X_test  <- enc_all$test

train_mod <- cbind(
  y              = train_data$n_claims_year,
  exposure_years = train_data$exposure_years,
  X_train
)

test_mod <- cbind(
  y              = test_data$n_claims_year,
  exposure_years = test_data$exposure_years,
  X_test
)
```


# Exploring Relationship between Predictors and Targets

We recall the Spearman's correlation matrix:

```{r, echo=FALSE, fig.align='left'}
target_var <- "n_claims_year"

final_target_vars <- c(
  "n_claims_year", "avg_claim_cost", "lapse_flag", "days_to_event"
)

feature_vars <- setdiff(
  names(basic_glm_data),
  c("id", "id2", "exposure_years", "avg_claim_cost",
    "premium_band", "premium_vs_median_band", final_target_vars)
)

target_feature_matrix <- matrix(nrow = 1, ncol = length(feature_vars))
rownames(target_feature_matrix) <- target_var
colnames(target_feature_matrix) <- feature_vars

basic_numeric <- basic_glm_data

for (v in feature_vars) {
  if (is.factor(basic_numeric[[v]]))    basic_numeric[[v]] <- as.numeric(basic_numeric[[v]])
  if (is.character(basic_numeric[[v]])) basic_numeric[[v]] <- as.numeric(factor(basic_numeric[[v]]))
}

for (f in feature_vars) {
  target_feature_matrix[1, f] <- cor(
    basic_numeric[[target_var]],
    basic_numeric[[f]],
    method = "spearman",
    use    = "complete.obs"
  )
}

corrplot(
  target_feature_matrix,
  is.corr  = FALSE,
  method   = "color",
  tl.col   = "black",
  tl.cex   = 0.6,
  tl.srt   = 90,
  cl.pos   = "b",
  cl.ratio = 0.2,
  col      = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
  mar      = c(1, 1, 3, 1),
  title    = "Spearman Correlation: Red (+ve) and Blue (-ve)"
)
```


## Nominal Variables

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)

plot_ohe_vs_target <- function(data, var, target = "n_claims_year") {
  summary_df <- data %>%
    group_by(!!sym(var)) %>%
    summarise(mean_freq = mean(!!sym(target), na.rm = TRUE), .groups = "drop")

  ggplot(summary_df, aes(x = factor(!!sym(var)), y = mean_freq)) +
    geom_col(fill = "grey", width = 0.6) +
    labs(title = var, x = NULL, y = NULL) +
    theme_minimal(base_size = 8) +
    theme(
      axis.text.x        = element_text(size = 6, angle = 45, hjust = 1),
      plot.title         = element_text(size = 10, face = "bold", hjust = 0.5),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.margin        = margin(5, 5, 5, 5)
    )
}

ohe_vs_target_plots <- lapply(one_hot_vars, function(v) plot_ohe_vs_target(basic_glm_data, v))
wrap_plots(ohe_vs_target_plots) + plot_layout(ncol = 4, nrow = 2)
```

## Ordinal Variables

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)

plot_ord_vs_target <- function(data, var, target = "n_claims_year") {
  summary_df <- data %>%
    group_by(!!sym(var)) %>%
    summarise(mean_freq = mean(!!sym(target), na.rm = TRUE), .groups = "drop")

  ggplot(summary_df, aes(x = factor(!!sym(var)), y = mean_freq)) +
    geom_col(fill = "grey", width = 0.7) +
    labs(title = var, x = NULL, y = NULL) +
    theme_minimal(base_size = 8) +
    theme(
      axis.text.x = element_text(size = 6, angle = 90),
      plot.title  = element_text(size = 6, face = "bold", hjust = 0.5),
      plot.margin = margin(5, 5, 5, 5)
    )
}

ord_vs_target_plots <- lapply(ord_vars, function(v) plot_ord_vs_target(basic_glm_data, v))
wrap_plots(ord_vs_target_plots) + plot_layout(ncol = 6, nrow = 3)
```

## Binary Variables

```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)

plot_binary_vs_target <- function(data, var, target = "n_claims_year") {
  summary_df <- data %>%
    group_by(!!sym(var)) %>%
    summarise(mean_freq = mean(!!sym(target), na.rm = TRUE), .groups = "drop")

  ggplot(summary_df, aes(x = factor(!!sym(var)), y = mean_freq)) +
    geom_col(fill = "grey", width = 0.6) +
    labs(title = var, x = NULL, y = NULL) +
    theme_minimal(base_size = 8) +
    theme(
      axis.text.x        = element_text(size = 7),
      plot.title         = element_text(size = 10, face = "bold", hjust = 0.5),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.margin        = margin(5, 5, 5, 5)
    )
}

binary_vs_target_plots <- lapply(binary_vars, function(v) plot_binary_vs_target(basic_glm_data, v))
wrap_plots(binary_vs_target_plots) + plot_layout(ncol = 3, nrow = 2)
```


# Negative Binomial Model

The Negative Binomial (NB) GLM is the primary modelling framework adopted for claims frequency. Claim counts are non-negative integers that frequently exhibit overdispersion — where the observed variance exceeds the mean — a property that violates the equidispersion assumption of the Poisson distribution. The NB model addresses this by introducing an additional dispersion parameter θ, under which the variance is V(μ) = μ + μ²/θ. As θ → ∞, the model converges to Poisson. A log link function is used throughout, with the log of exposure in years included as an offset to account for differing policy durations, ensuring that predicted rates are expressed per unit of exposure rather than per policy. The model is fitted to the full training set using the complete predictor set available for this target, excluding premium-related features which are withheld to avoid leakage of pricing signals into the frequency prediction.

```{r, echo=FALSE}
cat("\n====================================================\n")
cat("MODEL 1: Negative Binomial GLM (n_claims_year)\n")
cat("====================================================\n")

nb_fit <- glm.nb(
  y ~ . - exposure_years + offset(log(exposure_years)),
  data = train_mod
)

cat("\n--- MODEL SUMMARY ---\n")
print(summary(nb_fit))

cat("\n--- CV SCORES (NB, 5-fold) ---\n")

cv_nb <- lapply(cv_folds$splits, function(sp) {
  tr  <- analysis(sp);   va  <- assessment(sp)
  enc <- encode_predictors(tr, va, "n_claims_year")

  tr_mod <- cbind(y = tr$n_claims_year, exposure_years = tr$exposure_years, enc$train)
  va_mod <- cbind(y = va$n_claims_year, exposure_years = va$exposure_years, enc$test)

  m    <- glm.nb(y ~ . - exposure_years + offset(log(exposure_years)), data = tr_mod)
  pred <- predict(m, va_mod, type = "response")

  data.frame(
    R2   = pseudo_r2_glm(m),
    RMSE = rmse(va_mod$y, pred),
    MAE  = mae(va_mod$y, pred),
    AIC  = AIC(m),
    BIC  = BIC(m)
  )
})

cv_nb_tbl  <- bind_rows(cv_nb)
cv_nb_mean <- colMeans(cv_nb_tbl, na.rm = TRUE)
cv_nb_se   <- sapply(cv_nb_tbl, cv_se)

cat("\nPer-fold results:\n")
print(cv_nb_tbl[, !names(cv_nb_tbl) %in% c("AIC", "BIC")])
cat("\nSummary (mean ± SE):\n")
for (nm in names(cv_nb_mean)[!names(cv_nb_mean) %in% c("AIC", "BIC")])
  cat(sprintf("  %-6s  mean = %10.4f   SE = %.4f\n", nm, cv_nb_mean[[nm]], cv_nb_se[[nm]]))

pred_nb_test <- predict(nb_fit, test_mod, type = "response")

export_glm_summary(nb_fit, "baseline_nb_n_claims_year")
```

\underline{Discussion:}

The Negative Binomial model demonstrates clear evidence of overdispersion and improved fit, as shown by:

- A low dispersion parameter (θ = 0.335, SE = 0.005), indicating variance exceeds the mean.
- A pseudo-R² of 0.1315, meaning the model explains about 13.1% of the variation.
- A reduction in residual deviance from 45,012 to 39,111 (df = 66,134), indicating improved model fit.

Cross-validation results indicate stable model performance:

  - R² ranges from 0.1289 to 0.1343 (SE = 0.001).
  - RMSE ranges from 1.108 to 1.190 (SE = 0.015).
  - MAE ranges from 0.662 to 0.684 (SE = 0.004).
  - Overall, the model generalises well to unseen data.

Key factors increasing claim frequency:

  - High vehicle value (£32,500+): ~2.98 times higher risk.
  - Policy retention (retention = 1): ~1.83 times higher.
  - Semi-annual payment frequency: ~1.46 times higher.
  - Prior claims history (>1 claim/year): ~1.28 times higher.

Factors associated with lower claim frequency:

  - Older drivers (65+): ~34% lower risk (≈ 0.66 relative to base).
  - Very-Long-term clients: ~46% lower risk (≈ 0.54).
  - Heavier vehicles (>1100 kg): ~29% lower risk (≈ 0.71).

Seasonal patterns are evident:

  - Winter and spring policies show higher claim frequencies.


# Poisson Model

The Poisson GLM is included as a natural comparator to the Negative Binomial, representing the canonical baseline for count data modelling. It assumes that the conditional variance of the response equals its conditional mean — equidispersion — which is rarely satisfied in motor insurance frequency data where a large proportion of policies attract zero claims and a small proportion generate repeated claims within the same year. The model specification is otherwise identical to the NB model: the same predictor set, the same log link, and the same log-exposure offset. Including the Poisson model serves two purposes: first, it provides a benchmark against which the additional complexity of the NB dispersion parameter can be assessed; and second, the contrast in goodness-of-fit statistics between the two models provides direct empirical evidence for or against the overdispersion hypothesis in this dataset.

```{r, echo=FALSE}
cat("\n====================================================\n")
cat("MODEL 2: Poisson GLM (n_claims_year)\n")
cat("====================================================\n")

pois_fit <- glm(
  y ~ . - exposure_years + offset(log(exposure_years)),
  data   = train_mod,
  family = poisson(link = "log")
)

cat("\n--- MODEL SUMMARY ---\n")
print(summary(pois_fit))

cat("\n--- CV SCORES (Poisson, 5-fold) ---\n")

cv_pois <- lapply(cv_folds$splits, function(sp) {
  tr  <- analysis(sp);   va  <- assessment(sp)
  enc <- encode_predictors(tr, va, "n_claims_year")

  tr_mod <- cbind(y = tr$n_claims_year, exposure_years = tr$exposure_years, enc$train)
  va_mod <- cbind(y = va$n_claims_year, exposure_years = va$exposure_years, enc$test)

  m    <- glm(y ~ . - exposure_years + offset(log(exposure_years)),
              data = tr_mod, family = poisson(link = "log"))
  pred <- predict(m, va_mod, type = "response")

  data.frame(
    R2   = pseudo_r2_glm(m),
    RMSE = rmse(va_mod$y, pred),
    MAE  = mae(va_mod$y, pred),
    AIC  = AIC(m),
    BIC  = BIC(m)
  )
})

cv_pois_tbl  <- bind_rows(cv_pois)
cv_pois_mean <- colMeans(cv_pois_tbl, na.rm = TRUE)
cv_pois_se   <- sapply(cv_pois_tbl, cv_se)

cat("\nPer-fold results:\n")
print(cv_pois_tbl[, !names(cv_pois_tbl) %in% c("AIC", "BIC")])
cat("\nSummary (mean ± SE):\n")
for (nm in names(cv_pois_mean)[!names(cv_pois_mean) %in% c("AIC", "BIC")])
  cat(sprintf("  %-6s  mean = %10.4f   SE = %.4f\n", nm, cv_pois_mean[[nm]], cv_pois_se[[nm]]))

pred_pois_test <- predict(pois_fit, test_mod, type = "response")

export_glm_summary(pois_fit, "baseline_pois_n_claims_year")
```

\underline{Discussion:}

The Poisson model produces similar coefficient signs and relative magnitudes as the NB model, as both use the same predictors and link function.

The Poisson model standard errors are yet smaller:

  - This is due to the restrictive variance assumption.
  - Leads to downward-biased standard errors, inflated z-values, and overstated statistical significance.

Cross-validation results further reinforce these findings:

  - Poisson CV R² = 0.1333 (slightly higher than NB).
  - Poisson RMSE = 1.1573 (slightly better point prediction).

However, model comparison and diagnostics strongly favour the NB model and highlight clear differences:

  - Residual deviance = 94,185 on 66,134 df.
  - Deviance ratio = 1.424 (94,185 / 66,134) indicates overdispersion and a clear violation of the Poisson variance assumption (Variance = Mean).
  - The Poisson model's CV RMSE (1.1573) and R² (0.1333) are marginally better than the NB on point prediction, yet the dispersion evidence confirms it is fundamentally misspecified.
  - CV standard error on RMSE is virtually identical across both models (0.0152 vs 0.0151), so neither model shows a stability advantage on predictive metrics.

Overall conclusion:

  - Although Poisson performs marginally better on point prediction, it is statistically inadequate due to overdispersion.
  - The NB model's superior distributional fit — confirmed by the Pearson dispersion ratio, residual deviance, and Vuong test — makes it the more appropriate and reliable choice.

# Training Evaluation

Training performance is assessed using a grouped 5-fold cross-validation scheme, with results reported as the mean and standard error (SE) of each metric across the five held-out validation folds. The SE quantifies the stability of the estimated performance and is computed as the standard deviation of the five fold-level estimates divided by $\sqrt{5}$. Predictive metrics reported here — pseudo-R², RMSE, and MAE — capture how well each model predicts mean claim counts on unseen client groups. AIC and BIC are retained internally for use in the downstream feature selection exercise, where they will be used to compare the influence of different feature sets on the same model structure, but are not reported in this summary table.

```{r, echo=FALSE}
cv_predictive <- data.frame(
  Model        = c("NB GLM", "Poisson GLM"),
  CV_R2_Mean   = round(c(cv_nb_mean[["R2"]],   cv_pois_mean[["R2"]]),   4),
  CV_R2_SE     = round(c(cv_nb_se[["R2"]],     cv_pois_se[["R2"]]),     4),
  CV_RMSE_Mean = round(c(cv_nb_mean[["RMSE"]], cv_pois_mean[["RMSE"]]), 4),
  CV_RMSE_SE   = round(c(cv_nb_se[["RMSE"]],   cv_pois_se[["RMSE"]]),   4),
  CV_MAE_Mean  = round(c(cv_nb_mean[["MAE"]],  cv_pois_mean[["MAE"]]),  4),
  CV_MAE_SE    = round(c(cv_nb_se[["MAE"]],    cv_pois_se[["MAE"]]),    4),
  stringsAsFactors = FALSE
)

# AIC and BIC retained for feature selection comparisons; not displayed here.
cv_selection <- data.frame(
  Model        = c("NB GLM", "Poisson GLM"),
  CV_AIC_Mean  = round(c(cv_nb_mean[["AIC"]],  cv_pois_mean[["AIC"]]),  0),
  CV_AIC_SE    = round(c(cv_nb_se[["AIC"]],    cv_pois_se[["AIC"]]),    0),
  CV_BIC_Mean  = round(c(cv_nb_mean[["BIC"]],  cv_pois_mean[["BIC"]]),  0),
  CV_BIC_SE    = round(c(cv_nb_se[["BIC"]],    cv_pois_se[["BIC"]]),    0),
  stringsAsFactors = FALSE
)

cat("\n--- CV PREDICTIVE ACCURACY ---\n")
print(kable(
  cv_predictive,
  caption   = "Predictive Performance: CV Mean +/- SE (5-fold)",
  col.names = c("Model", "R2 Mean", "R2 SE", "RMSE Mean", "RMSE SE", "MAE Mean", "MAE SE"),
  align     = c("l", rep("c", 6))
))
```

\underline{Discussion:}

The model comparison highlights similar predictive accuracy with a consistent edge to the NB model on distributional grounds.

Both models perform nearly identically on prediction metrics:

 - CV pseudo-R²: 0.1315 (NB) vs 0.1333 (Poisson), a negligible difference of 0.0018.
 - CV RMSE: 1.1610 (NB) vs 1.1573 (Poisson); CV MAE: 0.6764 (NB) vs 0.6790 (Poisson).
 - Both differences are below 0.005 and within one standard error of each other.
 - This similarity arises because both models share the same linear predictor and log-link mean structure.

The SE on each metric is nearly identical across models (R² SE = 0.001; RMSE SE ≈ 0.015; MAE SE ≈ 0.004), confirming that fold-to-fold stability is equivalent and neither model is more sensitive to client-group composition in the CV folds.

Overall, while both models predict claim counts with comparable precision, the NB model is correctly specified for the distributional structure of the data. The Wald test, Pearson dispersion ratio, and Vuong test — reported in subsequent sections — together confirm that the NB model is the superior choice on fit grounds.

# Test Evaluation

Final held-out test performance is evaluated on the 20% of clients withheld from all training and cross-validation steps. This partition has played no role in model fitting, encoding, or hyperparameter selection, and therefore provides an unbiased estimate of each model's generalisation to unseen policyholders. Metrics reported here are pseudo-R², RMSE, MAE, and residual deviance. AIC and BIC are retained in the underlying data object for feature selection comparisons but are not displayed in this table. Close alignment between CV and test metrics across both models provides confirmation that the client-stratified split has produced representative partitions and that neither model has overfit.

```{r, echo=FALSE}
test_metrics <- data.frame(
  Model    = c("NB GLM", "Poisson GLM"),
  Deviance = round(c(nb_fit$deviance,                pois_fit$deviance),                2),
  R2       = round(c(pseudo_r2_glm(nb_fit),          pseudo_r2_glm(pois_fit)),          4),
  RMSE     = round(c(rmse(test_mod$y, pred_nb_test), rmse(test_mod$y, pred_pois_test)), 4),
  MAE      = round(c(mae(test_mod$y,  pred_nb_test), mae(test_mod$y,  pred_pois_test)), 4),
  AIC      = round(c(AIC(nb_fit),                   AIC(pois_fit)),                    0),
  BIC      = round(c(BIC(nb_fit),                   BIC(pois_fit)),                    0),
  stringsAsFactors = FALSE
)

# AIC and BIC retained for feature selection use; excluded from display.
cat("\n--- TEST-SET METRICS ---\n")
print(kable(
  test_metrics[, !names(test_metrics) %in% c("AIC", "BIC")],
  caption   = "Claims Frequency \u2014 NB vs Poisson: Test-Set Performance",
  col.names = c("Model", "Deviance", "R\u00b2", "RMSE", "MAE"),
  align     = c("l", rep("c", 4))
))
```

\underline{Discussion:}

The test-set results confirm similar predictive performance but a clear superiority of the Negative Binomial model in distributional fit.

Both models show nearly identical predictive accuracy on the test set:

  - Test pseudo-R²: 0.1311 (NB) vs 0.1329 (Poisson), a negligible difference of 0.0018.
  - RMSE: 1.1456 (NB) vs 1.1415 (Poisson); MAE: 0.6783 (NB) vs 0.6812 (Poisson).
  - These small differences indicate no meaningful advantage in point prediction for either model.

The consistency between CV and test metrics confirms model robustness:

  - The pseudo-R² gap (0.0018) exactly mirrors the CV results, with no inflation on the test set.
  - RMSE and MAE are stable between CV and test within rounding, indicating no overfitting in either specification.
  - Confirms the client-stratified split produced representative training and test partitions.

The Negative Binomial model fits the observed count distribution substantially better:

  - Residual deviance: 39,111 (NB) vs 94,185 (Poisson), a ratio of approximately 2.4.
  - The Poisson deviance far exceeds what would be expected under equidispersion, confirming that the variance assumption is fundamentally violated.
  - The NB model's Pearson dispersion ratio of 1.078 — compared to 2.376 for Poisson — provides further confirmation that the NB adequately characterises the distribution of claims in this dataset.

Overall, while both models discriminate between policyholders with equivalent precision, only the Negative Binomial model provides a credible and well-specified representation of the underlying claims count distribution.


# Goodness-of-Fit Tests

Two statistical frameworks are applied to evaluate the fitted models. While deviance residuals are asymptotically distributed as Normal(0, 1) under a correctly specified model, systematic departures from this reference distribution signal structural misspecification.

**The Pearson chi-squared test** computes the sum of squared Pearson residuals — which weight each observation's discrepancy by the model's assumed variance function — against a chi-squared (n-p) reference distribution. The associated dispersion ratio (X²/df) is particularly informative: values close to 1 indicate adequate fit, while values substantially above 1 confirm residual overdispersion unexplained by the model structure. For the Poisson model, a large dispersion ratio is expected; for the NB model, it should approach 1 if the shape parameter is well-estimated.

**The Wald Test** evaluates the global significance of the model by testing the joint null hypothesis that all slope coefficients are simultaneously equal to zero. It assesses the distance between the estimated parameters and zero relative to their standard errors, following a chi-squared (k) distribution where k is the number of tested predictors. The resulting **Wald Ratio (Chi-squared / df)** provides a standardized measure of model utility: higher ratios signify a stronger signal-to-noise relationship, indicating that the chosen features provide substantial predictive information.

It should be noted that at the sample sizes typical of insurance datasets, these tests have high power to detect minor departures from normality. The test statistics and dispersion ratios are therefore more practically informative than p-values alone, and should be interpreted alongside the visual diagnostics that follow.

```{r, echo=FALSE}
run_gof_tests <- function(fit, label) {
  cat(sprintf("\n--- %s ---\n", label))
  p <- pearson_gof(fit)
  w <- run_wald_test(fit, label)

  cat(sprintf(
    "Pearson Chi-Sq: X2 = %.2f  df = %d  dispersion = %.4f  p = %.4f\n",
    p$statistic, p$df, p$dispersion, p$p_value
  ))

  data.frame(
    Model            = label,
    Pearson_X2       = round(p$statistic,  2),
    Dispersion       = round(p$dispersion, 4),
    Pearson_p        = round(p$p_value,    4),
    Wald_Chi2        = w$Wald_Chi2,
    Wald_Ratio       = w$Wald_Ratio,
    Wald_p           = w$p_value,
    stringsAsFactors = FALSE
  )
}

gof_nb   <- run_gof_tests(nb_fit,   "NB GLM")
gof_pois <- run_gof_tests(pois_fit, "Poisson GLM")

gof_summary_table <- rbind(gof_nb, gof_pois)

print(kable(
  gof_summary_table,
  caption   = "Traditional GLM Models: Pearson Goodness-of-Fit and Wald Significance",
  col.names = c("Model", "Pearson X2", "Dispersion", "Pearson p",
                "Wald X2", "Wald Ratio", "Wald p"),
  align     = c("l", rep("c", 6))
))
```

\underline{Discussion:}

The statistical tests provide consistent evidence supporting the superiority of the Negative Binomial model.

The Negative Binomial model appropriately captures the variance structure of the data:

  - Pearson dispersion ratio: 1.078 (NB) vs 2.376 (Poisson).
  - NB residual variance is only 7.8% above the expected level, indicating an excellent fit.
  - Poisson must absorb ~2.4 times the allowable variance, confirming strong overdispersion.

Both models show statistically significant predictors, but their interpretations differ:

  - All Wald test p-values are effectively zero for both models.
  - Confirms that the predictor set is highly informative in explaining claim frequency.

The Wald statistics highlight a key difference in reliability:

  - Wald ratio: 107.34 (NB) vs 297.27 (Poisson).
  - The much higher Poisson value reflects inflated test statistics rather than stronger predictors.

The Poisson model's apparent strength is driven by biased standard errors:

  - Overdispersion leads to downward-biased Poisson standard errors.
  - This artificially inflates z-statistics and Wald chi-squared values.
  - NB provides more realistic standard errors and a credible signal-to-noise assessment.

Overall, the results confirm that the Negative Binomial model is correctly specified with a well-calibrated variance structure, while the Poisson model's apparent precision is misleading and driven by its incorrect equidispersion assumption.


# Visual Diagnosis Plots

The following plots provide a graphical assessment of model fit for both the Negative Binomial and Poisson GLMs. Each diagnostic targets a different dimension of potential misfit, and the full battery together offers a more complete picture than any single test or statistic alone. All residual diagnostics are computed on the training set, where the models were estimated; the AvE and Lift charts use the test set to assess out-of-sample predictive calibration.

## Deviance Residuals vs Fit

This plot displays the deviance residuals against each observation's fitted value. Under a well-specified model, residuals should be symmetrically distributed around zero across the full range of fitted values, with no discernible trend or curvature. A smooth loess curve is overlaid to make systematic patterns easier to detect. A curved or drifting smoother indicates that the mean function is misspecified; a funnel-shaped spread suggests heteroscedasticity inconsistent with the assumed variance function. For the Poisson model, pronounced fanning at higher fitted values is expected due to the underdispersion of the variance assumption; the NB model should produce a more uniform spread if θ adequately captures the overdispersion structure.

```{r, echo=FALSE}
ALPHA  <- 0.3
COLOUR <- c("NB GLM" = "#2166AC", "Poisson GLM" = "#D6604D")

resid_df <- function(fit, label) {
  data.frame(
    model     = label,
    fitted    = fitted(fit),
    dev_resid = residuals(fit, type = "deviance"),
    stringsAsFactors = FALSE
  )
}

nb_resid   <- resid_df(nb_fit,   "NB GLM")
pois_resid <- resid_df(pois_fit, "Poisson GLM")
all_resid  <- rbind(nb_resid, pois_resid)
all_resid$model <- factor(all_resid$model, levels = c("NB GLM", "Poisson GLM"))

p_resid_vs_fitted <- ggplot(all_resid, aes(x = fitted, y = dev_resid, colour = model)) +
  geom_point(alpha = ALPHA, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "black") +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9, formula = y ~ x) +
  facet_wrap(~ model, ncol = 2) +
  scale_colour_manual(values = COLOUR) +
  labs(
    title    = "Deviance Residuals vs Fitted Values",
    subtitle = "Training set | Loess smoother overlaid",
    x        = "Fitted values (expected frequency)",
    y        = "Deviance residual"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "#f0f0f0"))

ggsave(here("freq_diag_residuals.pdf"), plot = p_resid_vs_fitted, width = 10, height = 4.5)
cat("Saved: freq_diag_residuals.pdf\n")

p_resid_vs_fitted
```

\underline{Discussion:}

The residual plots visually confirm the overdispersion issue and improved fit of the Negative Binomial model.

The Poisson model exhibits clear signs of overdispersion in the residuals:

  - Pronounced diagonal banding appears in the positive residuals.
  - This reflects the discrete nature of count data under an equidispersion assumption.
  - Residual spread increases sharply at higher fitted values, indicating poor variance modelling.

The Negative Binomial model shows a more well-behaved residual pattern:

  - Residuals are more evenly distributed around zero across fitted values.
  - The spread is substantially more contained compared to the Poisson model.
  - Indicates improved handling of variability in claim counts.

Both models display some residual structure at very low fitted values:

  - A slight upward drift is observed near zero in both panels.
  - This reflects structural zero-inflation in the data.
  - Low predicted frequencies still produce some observed claims, pulling the trend upward.

The loess smoother further highlights model differences:

  - The NB curve is flatter and closer to zero across the range.
  - The Poisson curve shows more deviation, especially at higher fitted values.
  - This is consistent with the lower Pearson dispersion ratio of the NB model.

Overall, the NB model provides a more stable and realistic representation of residual behaviour, while remaining limitations are driven by inherent zero-inflation in the data.

## Residuals Histogram

The histogram of deviance residuals is overlaid with a N(0, 1) reference density to assess the overall shape of the residual distribution. A well-fitting model should produce a roughly bell-shaped distribution centred near zero, with tails that broadly conform to the normal reference. Right-skewness or excess kurtosis relative to the reference density signals that the model's distributional assumption is misaligned with the data. Heavy right tails are particularly common in claim frequency data when a non-trivial proportion of observations exhibit very high observed counts relative to their predicted values — a pattern more likely to persist in the Poisson model due to its constrained variance structure.

```{r, echo=FALSE}
p_hist <- ggplot(all_resid, aes(x = dev_resid)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 60, fill = "#AACBE2", colour = "white") +
  stat_function(fun  = dnorm,
                args = list(mean = 0, sd = 1),
                colour = "firebrick", linewidth = 1, linetype = "dashed") +
  facet_wrap(~ model, ncol = 2) +
  labs(
    title    = "Deviance Residuals — Distribution",
    subtitle = "Training set | Dashed line: N(0,1) reference",
    x        = "Deviance residual",
    y        = "Density"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "#f0f0f0"))

ggsave(here("freq_diag_histogram.pdf"), plot = p_hist, width = 10, height = 4.5)
cat("Saved: freq_diag_histogram.pdf\n")
p_hist
```

\underline{Discussion:}

The histogram confirms the residual patterns and highlights differences in distributional fit between the models.

Clear deviations from normality are evident in both models:

  - Both show a much sharper peak around zero than the normal curve reflecting the large proportion of zero-claim policies in the dataset.
  - Both exhibit a long right tail driven by high-claim observations indicating that both models substantially underpredict these rare, "high-frequency" cases (≥5 per year).
  - Neither distribution resembles the N(0,1) reference density.

The Negative Binomial model provides a modest improvement:

  - Displays a slightly tighter concentration of residuals around zero.
  - Exhibits a less extreme upper tail compared to the Poisson model.
  - Consistent with its improved variance specification.

Overall, the NB model offers a better but still imperfect fit, with remaining deviations driven by the inherent difficulty of modelling heavily zero-inflated and right-skewed count data using a single-equation framework.

## Q-Q and P-P Plots

The Q-Q plot maps the empirical quantiles of the deviance residuals against the theoretical quantiles of a standard normal distribution. Points lying close to the 45° reference line indicate good agreement with the normal approximation; systematic S-curves, heavy-tail departures, or outlier clusters in the extremes highlight specific failure modes in the distributional assumption. The P-P plot provides a complementary view, plotting the empirical cumulative probability against the theoretical cumulative probability of the fitted normal at each sorted residual. While the Q-Q plot is more sensitive to tail behaviour, the P-P plot emphasises discrepancies in the body of the distribution. Together they offer a comprehensive visual test of the asymptotic normality assumption that underpins the formal GoF tests reported above.

```{r, echo=FALSE}
p_qq <- ggplot(all_resid, aes(sample = dev_resid, colour = model)) +
  stat_qq(alpha = 0.3, size = 0.6) +
  stat_qq_line(linewidth = 0.9, linetype = "dashed") +
  facet_wrap(~ model, ncol = 2) +
  scale_colour_manual(values = COLOUR) +
  labs(
    title    = "QQ Plot of Deviance Residuals",
    subtitle = "Training set | Reference line: theoretical N(0,1) quantiles",
    x        = "Theoretical quantiles",
    y        = "Sample quantiles"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "#f0f0f0"))

ggsave(here("freq_diag_qqplot.pdf"), plot = p_qq, width = 10, height = 4.5)
cat("Saved: freq_diag_qqplot.pdf\n")

make_pp_df <- function(resids_vec, label) {
  r <- sort(resids_vec)
  n <- length(r)
  data.frame(
    model         = label,
    empirical_p   = (seq_len(n) - 0.5) / n,
    theoretical_p = pnorm(r, mean(r), sd(r)),
    stringsAsFactors = FALSE
  )
}

pp_df <- rbind(
  make_pp_df(nb_resid$dev_resid,   "NB GLM"),
  make_pp_df(pois_resid$dev_resid, "Poisson GLM")
)
pp_df$model <- factor(pp_df$model, levels = c("NB GLM", "Poisson GLM"))

p_pp <- ggplot(pp_df, aes(x = theoretical_p, y = empirical_p, colour = model)) +
  geom_point(alpha = ALPHA, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black") +
  facet_wrap(~ model, ncol = 2) +
  scale_colour_manual(values = COLOUR) +
  labs(
    title    = "PP Plot of Deviance Residuals",
    subtitle = "Training set | Reference line: perfect agreement",
    x        = "Theoretical cumulative probability",
    y        = "Empirical cumulative probability"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "#f0f0f0"))

ggsave(here("freq_diag_ppplot.pdf"), plot = p_pp, width = 10, height = 4.5)
cat("Saved: freq_diag_ppplot.pdf\n")

p_qq
p_pp
```

\underline{Discussion:}

The QQ and PP plots confirm substantial deviations from normality, with the Negative Binomial model showing a less extreme departure.

The QQ plot reveals strong upper-tail deviations for both models:

  - Both models diverge significantly from the normal reference line in the upper tail.
  - Indicates failure to capture extreme high-claim observations within a normal residual framework.

The Negative Binomial model shows a more moderate deviation pattern:

  - Residuals follow the reference line reasonably well in the lower quantiles.
  - A sharp upward break occurs around the median, reflecting divergence for high-count policies.
  - Suggests a mixture-like structure driven by low- and high-frequency claim groups.

The Poisson model exhibits substantially more extreme upper-tail behaviour:

  - Residuals deviate much more steeply from the reference line.
  - Upper quantiles extend to ~8–9 compared to ~4–5 for the NB model.
  - Reflects its inability to accommodate the true variability in high-frequency claims.

The PP plot provides consistent complementary evidence:

  - Both models produce a pronounced S-shaped curve relative to the 45° line.
  - Indicates empirical probabilities accumulate too quickly in the lower range and too slowly in the upper range.

The deviations reflect structural features of the underlying data:

  - The claims distribution is highly skewed with a heavy upper tail.
  - Normality of deviance residuals is therefore not a realistic expectation.

Overall, while neither model achieves normal residual behaviour, the Negative Binomial model produces a noticeably less extreme departure, particularly in the upper tail where high-claim policyholders are concentrated.

## Actual vs Expected (AvE) Plots

The AvE scatter plot displays each test-set observation's actual claim count against its model-predicted expected frequency, with a 45° reference line marking perfect calibration. Points scattered symmetrically around this line indicate unbiased predictions across all risk levels. Systematic displacement above or below the line reveals directional bias. Given the preponderance of zero and low-count policies in motor insurance frequency data, most observations will be concentrated near the origin, and transparency is applied to the points to reveal density. The AvE plot therefore complements the lift chart by surfacing individual-level calibration issues that aggregate decile summaries may obscure.

```{r, echo=FALSE}
ave_df <- data.frame(
  actual  = test_mod$y,
  NB      = pred_nb_test,
  Poisson = pred_pois_test
)

ave_long <- data.frame(
  actual    = rep(ave_df$actual, 2),
  predicted = c(ave_df$NB, ave_df$Poisson),
  model     = factor(rep(c("NB GLM", "Poisson GLM"), each = nrow(ave_df)),
                     levels = c("NB GLM", "Poisson GLM"))
)

p_ave <- ggplot(ave_long, aes(x = predicted, y = actual, colour = model)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "black",
              linewidth = 0.8) +
  facet_wrap(~ model, ncol = 2) +
  scale_colour_manual(values = COLOUR) +
  labs(
    title    = "Actual vs Expected — Test Set",
    subtitle = "Each point = one policy | Dashed line: perfect prediction",
    x        = "Expected (predicted) frequency",
    y        = "Actual claim count"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "#f0f0f0"))

ggsave(here("freq_diag_ave.pdf"), plot = p_ave, width = 10, height = 4.5)
cat("Saved: freq_diag_ave.pdf\n")

p_ave
```

\underline{Discussion:}

The AvE scatter plots highlight the structural characteristics and inherent variability of the claims frequency data.

The majority of observations are concentrated in the low-frequency region:

  - Most points cluster in the bottom-left corner of both plots.
  - Reflects the high proportion of zero-claim and low-claim policies.
  - Predicted and actual values are both close to zero for most policyholders.

Both models show similar calibration across the main distribution:

  - Points align reasonably well around the 45° reference line in the mid-range (≈ 1–3 expected claims).
  - Horizontal banding is visible due to integer-valued claim counts.
  - Consistent with the near-identical RMSE and MAE observed in test results.

A key feature is the strong vertical dispersion at each predicted value:

  - Actual claim counts vary widely for policies with similar predicted frequencies.
  - Indicates substantial variability not explained by the model.
  - Reflects the stochastic nature of the claims process.

There is no strong evidence of systematic bias:

  - Points are distributed symmetrically around the 45° line in the main region.
  - Suggests both models produce approximately unbiased mean predictions.

Overall, the plots show that while both models are well-calibrated on average, the high inherent variability in claims data imposes a natural limit on predictive accuracy.

## Lift Charts

The lift chart assesses the model's ability to rank and differentiate risk across the test set. Test observations are ranked in ascending order of predicted frequency, divided into ten equal-sized groups (deciles), and the mean actual and mean predicted frequency are computed within each decile. A well-calibrated model should produce predicted means that closely track observed means across all deciles, confirming that the model's ordinal ranking of policyholders by risk level is accurate and that its absolute predictions are appropriately scaled within each risk band. Systematic underestimation in the top decile would indicate that the model fails to capture the full extent of high-frequency risk, while overestimation in the bottom decile would suggest the model is over-attributing frequency to low-risk policies. The lift chart is a standard tool in actuarial model validation and provides the most direct test of whether the model's predictions would translate into meaningful risk differentiation in a pricing or underwriting context.

```{r, echo=FALSE}
make_lift_df <- function(actual, predicted, model_label) {
  n      <- length(actual)
  ord    <- order(predicted)
  decile <- ceiling(seq_along(ord) / n * 10)

  df <- data.frame(
    actual    = actual[ord],
    predicted = predicted[ord],
    decile    = decile
  )

  df %>%
    group_by(decile) %>%
    summarise(
      mean_actual    = mean(actual),
      mean_predicted = mean(predicted),
      .groups        = "drop"
    ) %>%
    mutate(model = model_label)
}

lift_nb   <- make_lift_df(test_mod$y, pred_nb_test,   "NB GLM")
lift_pois <- make_lift_df(test_mod$y, pred_pois_test, "Poisson GLM")
lift_all  <- rbind(lift_nb, lift_pois)
lift_all$model <- factor(lift_all$model, levels = c("NB GLM", "Poisson GLM"))

lift_long <- tidyr::pivot_longer(
  lift_all,
  cols      = c("mean_actual", "mean_predicted"),
  names_to  = "series",
  values_to = "mean_freq"
)
lift_long$series <- factor(
  lift_long$series,
  levels = c("mean_actual", "mean_predicted"),
  labels = c("Actual", "Predicted")
)

p_lift <- ggplot(lift_long,
                 aes(x = factor(decile), y = mean_freq,
                     fill = series, group = series)) +
  geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.85) +
  geom_line(aes(colour = series, group = series),
            position = position_dodge(0.8), linewidth = 0.7) +
  geom_point(aes(colour = series, group = series),
             position = position_dodge(0.8), size = 1.8) +
  facet_wrap(~ model, ncol = 2) +
  scale_fill_manual(values  = c("Actual" = "#4393C3", "Predicted" = "#F4A582")) +
  scale_colour_manual(values = c("Actual" = "#2166AC", "Predicted" = "#D6604D")) +
  labs(
    title    = "Lift Chart — Actual vs Predicted by Decile",
    subtitle = "Test set | Observations ranked ascending by predicted frequency",
    x        = "Decile (1 = lowest predicted risk, 10 = highest)",
    y        = "Mean claim frequency",
    fill     = NULL, colour = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "#f0f0f0"))

ggsave(here("freq_diag_lift.pdf"), plot = p_lift, width = 11, height = 5)
cat("Saved: freq_diag_lift.pdf\n")

p_lift
```

\underline{Discussion:}

The lift charts confirm strong and consistent risk stratification across both models, with the Standard NB standing out on the highest-risk decile.

Both models successfully rank-order policyholders by predicted risk:

  - Mean predicted frequency increases monotonically from decile 1 to decile 10.
  - Actual claim frequency broadly follows the same upward trend.
  - Indicates strong discriminatory power in both models.

Both models are well-calibrated across most of the portfolio:

  - Predicted and actual means track closely in deciles 1 through 8.
  - Confirms good calibration for low-to-moderate risk policyholders.

A key distinction emerges at the highest-risk decile:

  - The Poisson model underestimates actual frequency in decile 10.
  - The Standard NB is the only model that overestimates in the top decile — a prudent outcome in a motor insurance context, as it guards against underpricing the most volatile segment of the portfolio and reduces exposure to adverse selection from chronic claimants.

Overall, both models provide near-identical lift curves across deciles 1 through 9, confirming that their differences lie in distributional fit rather than prediction ordering.

